Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint

### 2019-10-21 analysis/VOCC/1-split-biomass-by-maturity.Rmd
tweaks to depth-only biomass models

---
title: "Split `r paste(params$species)` biomass by maturity class"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Curlfin Sole"
    region: "All synoptic surveys"
    #region: "Both odd year surveys"
---

>>>>

---
title: "Split `r paste(params$species)` biomass by maturity class"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Cod"
    region: "All synoptic surveys"
    #region: "Both odd year surveys"
---

### 2019-10-21 analysis/VOCC/2-biomass-depth-only-model.Rmd
tweaks to depth-only biomass models

---
title: "Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Copper Rockfish"
    region: "All synoptic surveys"
    covariates: "+trawled+as.factor(ssid)" # additional fixed-effect covariates
    covs: "-trawled-ssid" # covariate string used to name model rds 
---

@@

```{r params}
species <- params$species
region <- params$region
covariates <- params$covariates
covs <- params$covs
priors <- FALSE

paste("region =", region)
paste("model covariates =", covariates)
paste("model label =", covs)
paste("priors =", priors)
```

@@

# scale predictors before filtering to ensure mean and SD are global
data <- data %>% mutate(raw_depth = depth, depth = log(raw_depth))
data <- gfranges::scale_predictors(data, # predictors = c(quo(depth)))
  predictors = c(quo(depth), quo(mixed), quo(muddy), quo(sandy), quo(rocky), quo(any_rock))
)
data <- data %>%
  mutate(depth = raw_depth) %>%
  filter(ssid %in% model_ssid)
```

@@

Run climate independent sdmTMB model
```{r}
tictoc::tic()
if (any(names(data) == "adult_density")) {
  adult_formula <- as.formula(paste(
    "adult_density ~ 0 + as.factor(year)", covariates, ""
  ))

@@

  imm_biomass <- sdmTMB::sdmTMB(data,
    imm_formula,
    time_varying = ~ 0 + depth_scaled + depth_scaled2, #+ trawled
    time = "year", spde = spde,
    family = tweedie(link = "log"),
    ar1_fields = FALSE,
    include_spatial = TRUE,
    enable_priors = priors,
    control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
    silent = FALSE
  )

@@

  saveRDS(total_biomass, file = paste0(
    "data/", spp,
    "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))
}
tictoc::toc()
```

Make prediction grids for all surveys and years, if not done before.
```{r, message=FALSE, warning=FALSE}
try(nd_all <- readRDS("data/nd_all_synoptic.rds"))

@@

  # create grids for each ssid separately
  ssid <- 1
  survey_abbrev <- "SYN QCS"
  nd_1 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)
  unique(nd_1$year)

@@

  newcov <- readRDS("data/new_covariates.rds") %>% dplyr::select(X, Y, trawled, muddy, rocky, mixed, any_rock)
  nd_all <- left_join(nd_all, newcov, by = c("X", "Y"))

  # nd <- nd %>% dplyr::mutate(shallow = ifelse(depth > 35, 0, 1))

@@

  imm_biomass <- readRDS(paste0("data/", spp,
    "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))
  imm_predictions <- predict(imm_biomass, newdata = nd)
  saveRDS(imm_predictions, file = paste0("data/", spp,
    "/predictions-", spp, covs, "-", ssid_string, "-imm-biomass-prior-", priors, ".rds"))

} else {
    nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
    nd <- nd_all %>% filter(ssid %in% model_ssid)
    nd <- na.omit(nd)
    nd$year <- as.integer(nd$year)

    total_biomass <- readRDS(paste0("data/", spp,
      "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"))
    predictions <- predict(total_biomass, newdata = nd)
    saveRDS(predictions, file = paste0("data/", spp,
      "/predictions-", spp, covs, "-", ssid_string, "-total-biomass-prior-", priors, ".rds"))
}
```

>>>>

---
title: "Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Petrale Sole"
    region: "All synoptic surveys"
    knots: 600
    covariates: "+trawled+as.factor(ssid)" # additional fixed-effect covariates
    covs: "-trawled-ssid" # covariate string used to name model rds 
    update_model: TRUE
---

@@

```{r params}
species <- params$species
region <- params$region
covariates <- params$covariates
covs <- params$covs
knots <- params$knots
priors <- FALSE

paste("region =", region)
paste("model covariates =", covariates)
paste("model label =", covs)
paste("priors =", priors)
paste("knots =", knots)
```

@@

# scale predictors before filtering to ensure mean and SD are global
data <- data %>% mutate(raw_depth = depth, depth = log(raw_depth))
data <- gfranges::scale_predictors(data, # predictors = c(quo(depth)))
  predictors = c(quo(depth), quo(mixed), quo(muddy), quo(sandy), quo(rocky), quo(any_rock))
)
data <- data %>%
  mutate(depth = raw_depth) %>%
  filter(ssid %in% model_ssid) %>% 
  filter(year < 2019)
```

@@

Run climate independent sdmTMB model
```{r }
if (params$update_model) {
tictoc::tic()
if (any(names(data) == "adult_density")) {
  adult_formula <- as.formula(paste(
    "adult_density ~ 0 + as.factor(year)", covariates, ""
  ))

@@

  imm_biomass <- sdmTMB::sdmTMB(data,
    imm_formula,
    time_varying = ~ 0 + depth_scaled + depth_scaled2, #+ trawled
    time = "year", spde = spde,
    family = tweedie(link = "log"),
    ar1_fields = TRUE, # changed to TRUE on Oct 17 2019
    include_spatial = TRUE,
    enable_priors = priors,
    control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
    silent = FALSE
  )

@@

  saveRDS(total_biomass, file = paste0(
    "data/", spp,
    "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))
}
tictoc::toc()
}
```

Make prediction grids for all surveys and years, if not done before.
```{r, eval=FALSE, message=FALSE, warning=FALSE}
try(nd_all <- readRDS("data/nd_all_synoptic.rds"))

@@

  # create grids for each ssid separately
  ssid <- 1
  survey_abbrev <- "SYN QCS"
  nd_1 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)

  unique(nd_1$year)

@@

  newcov <- readRDS("data/new_covariates.rds") %>% dplyr::select(X, Y, trawled, muddy, rocky, mixed, any_rock)
  nd_all <- left_join(nd_all, newcov, by = c("X", "Y"))
  nd_all$trawled[is.na(nd_all$trawled)] <- 0 # correct missing values for small area on edge of QCS grid
  # nd <- nd %>% dplyr::mutate(shallow = ifelse(depth > 35, 0, 1))

@@

  imm_biomass <- readRDS(paste0("data/", spp,
    "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))

    nd <- nd_all %>%
    filter(ssid %in% model_ssid) %>%
    filter(year %in% unique(imm_biomass$data$year))
  nd <- na.omit(nd)
  nd$year <- as.integer(nd$year)

  imm_predictions <- predict(imm_biomass, newdata = nd)
  saveRDS(imm_predictions, file = paste0("data/", spp,
    "/predictions-", spp, covs, "-", ssid_string, "-imm-biomass-prior-", priors, ".rds"))

    total_biomass <- readRDS(paste0("data/", spp,
      "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"))

    nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
    nd <- nd_all %>% 
      filter(ssid %in% model_ssid) %>%
      filter(year %in% unique(total_biomass$data$year))
    nd <- na.omit(nd)
    nd$year <- as.integer(nd$year)

    predictions <- predict(total_biomass, newdata = nd)
    saveRDS(predictions, file = paste0("data/", spp,
      "/predictions-", spp, covs, "-", ssid_string, "-total-biomass-prior-", priors, ".rds"))
}
```
